---
title: "Assignment 04 - Data Science for Public Policy"
author: "Eric Kitaif - ek910"
output: 
  html_document
urlcolor: blue
---

```{r include = FALSE}
def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::opts_chunk$set(cache = TRUE)
knitr::knit_hooks$set(
  chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(options$size != "normalsize", paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize"), x)
  }
)
# knitr::knit_hooks$set(inline = function(x) {
#   prettyNum(round(x, 2), big.mark = ",")
# })
options(scipen=999)
```


\begin{center}
\Huge{PPOL 670 | Assignment 04}

\Huge{Data Visualization, Markdown, and Git}
\end{center}

\vspace{0.1in}

## Setup
```{r setup, message=FALSE, cache=FALSE}
library(tigris)
library(sf)
library(tidyverse)
library(tidync)
library(scales)
```

## Links

[GitHub Repo](https://github.com/EricKit/assignment04)

The [Global Dataset of Historical Yield](https://doi.pangaea.de/10.1594/PANGAEA.909132) offers yield values for different crops from 1981-2016.

The [NClimGrid](https://www.ncei.noaa.gov/access/metadata/landing-page/bin/iso?id=gov.noaa.ncdc:C00332) data from NOAA provides United States temperature and precipitation date geographically from 1895 to present.

[AQUASTAT](http://www.fao.org/aquastat/statistics/query/results.html) contains a webquery which can output a csv file. There are plenty of variables to select: of note precipitation, water use, and renewable water resources from 1958 to 2020.

## Data Citations

NOAA Data: Vose, Russell S., Applequist, Scott, Squires, Mike, Durre, Imke, Menne, Matthew J., Williams, Claude N. Jr., Fenimore, Chris, Gleason, Karin, and Arndt, Derek (2014): NOAA Monthly U.S. Climate Gridded Dataset (NClimGrid), Version 1.

Wheat Data: Iizumi, Toshichika (2019): Global dataset of historical yields v1.2 and v1.3 aligned version. PANGAEA, https://doi.org/10.1594/PANGAEA.909132, Supplement to: Iizumi, Toshichika; Sakai, T (2020): The global dataset of historical yields for major crops 1981–2016. Scientific Data, 7(1), https://doi.org/10.1038/s41597-020-0433-7

Aquastat: FAO 2021. AQUASTAT Database. AQUASTAT Website  accessed on [12/10/2021 23:21].

## Retrieve Data

### Read in wheat data

```{r read_wheat, cache=TRUE, warning=FALSE}
file_list <- list.files("data/gdhy_v1/wheat/", pattern="nc4")

rm(wheat_data)

## Loop through all the files and merge them into one tibble

for(file_name in file_list) {
  path <- paste("data/gdhy_v1/wheat/", file_name, sep="")
  file <- tidync(path)
  year <- parse_number(file_name)
  
  new_data <- file %>%
    
    # Remove lats and lons that are not in the other data sets and convert lon
    # from 0-360 to -180 to 180
    hyper_filter(lat = lat >= 24.6 & lat <= 49.4,
                 lon = lon >= (-125 %% 360) & lon <= (-67 %% 360)) %>% 
    hyper_tibble() %>% 
    mutate(lon = ((lon + 180) %% 360) - 180) %>% 
    rename(yield = var) %>% 
    add_column(year = year)
  
  if (exists("wheat_data")){
    wheat_data <- bind_rows(wheat_data, new_data)
  } else {
    wheat_data <- new_data
  }
}
```

### Read in temperature data from NOAA

```{r read_noaa, cache=TRUE}
path <- "data/noaa/nclimgrid_tavg.nc"
file <- tidync(path)

# Time is in monthly increments, but the units are days since 1800-01-01.
# Here we pull the data for July from 1981 to 2016.

tavg_data <- file %>%
  hyper_filter(time = time >= 66291 & time <= 79106) %>%
  hyper_tibble() %>%
  filter(
    
    # Only include data that is near Wheat data
    # Cannot filter non-continuous data, so I have to do it after making the tibble.
    (lat - 0.25) %% 0.5 < 0.05,
    (lon + 0.25) %% 0.5 < 0.05,
    
    # Finds only the July month
    # The 365 * 4 - 45 accounts for leap years starting at 1981.
    time %% 365 - (floor(time / (365 * 4)) - 45) >= 226 &
      time %% 365 - (floor(time / (365 * 4)) - 45) <= 256
  ) %>%
  mutate(
    lat = lat - (lat %% 0.25),
    lon = lon - (lon %% 0.25),
    year = 1800 + floor(time / 365)
  ) %>%
  select(-time)

```

## Merge data and prepare datasets

```{r merge, cache=TRUE, message=FALSE}

data <- left_join(tavg_data, wheat_data)

data_sf = st_as_sf(data, coords=c("lon", "lat"))

temp_1982 <- file %>%
  hyper_filter(time = time >= 66656 & time <= 66686) %>%
  hyper_tibble() %>% 
  select(-time) %>% 
  add_column(year = 1982)

temp_2015 <- file %>%
  hyper_filter(time = time >= 78709 & time <= 78739) %>%
  hyper_tibble() %>% 
  select(-time) %>% 
  add_column(year = 2015)

tavg_detailed_sf <- bind_rows(temp_1982, temp_2015) %>% 
  st_as_sf(coords=c("lon", "lat"))

states <- states(cb = TRUE, progress_bar = FALSE) %>% 
  filter(STUSPS != "HI",
         STUSPS != "AK", 
         STUSPS != "VI",
         STUSPS != "AS",
         STUSPS != "MP",
         STUSPS != "GU",
         STUSPS != "PR")
```

## Visualization 1

The first graphs show the average temperature of America in July of 1982 and 2015. This data is colored based on the optimal temperature to grow wheat. Blue is too cold, red is too hot, yellow is optimal. This graph gives a good picture of where the hot and cold regions of America are with respect to wheat growth. A few areas stand out to me. In the Northwest region there are areas which became more suitable to wheat in 2015. California and the Southeast have fewer areas that were good for wheat in 2015. Interestingly, Kansas appears to have gained more area to farm in 2015.

```{r vis_1, cache=TRUE}
data_sf %>% 
  filter(year == 1982 | year == 2015) %>% 
  # Ideal wheat growing temperature
  st_set_crs(value = 4326) %>%
  ggplot() + 
  geom_sf(aes(color = tavg)) + 
  scale_color_gradientn(colors = c("blue", "blue","yellow", "yellow", "red", "red"),
                        values=(rescale(c(6, 11, 12.22, 24, 25, 37)))) + 
  geom_sf(data = states, fill = NA) + 
  facet_wrap(~ year, nrow = 2) + 
  labs(
      title = "Ideal Wheat Growth",
      subtitle = "The yellow region is the ideal temperature for wheat",
      color="\u00B0C")
```

## Visualization 2

Based on the first graph, I wanted to see which areas of America have changed temperature obtained from an average that better shows the comparison. The values are the average July temperature from 2011-2016 minus the average July temperature from 1981-1987. As you can see, throughout most of America the average temperature did go up. However, the average temperature from Kansas went down, which is the same area where we saw wheat could grow better in the previous graph.

```{r vis_2, cache=TRUE}
first_5 <- data %>% 
  filter(year <= 1987) %>% 
  group_by(lat, lon) %>%
  summarize(first_5_avg = mean(tavg))

last_5 <- data %>% 
  filter(year >= 2011) %>% 
  group_by(lat, lon) %>% 
  summarize(last_5_avg = mean(tavg))

temp_diff <- left_join(first_5, last_5, by = c("lat", "lon")) %>% 
  mutate(diff = last_5_avg - first_5_avg)

temp_diff %>% 
  st_as_sf(coords=c("lon", "lat")) %>% 
  st_set_crs(value = 4326) %>%
  ggplot() + 
  geom_sf(aes(color = diff)) +
  scale_color_gradient2(midpoint = 0, low = "blue", mid = "white", high = "red") +
  geom_sf(data = states, fill = NA) + 
  labs(
      title = "Temperature Change",
      subtitle = "Five year average around 2016 compared to  1981",
      color="\u00B0C Change")
```

## Visualization 3

The weighted average location for the production of wheat are shown for each year. The hypothesis to test was if the area where wheat is being grown is in fact moving. To attempt to answer this, this figure shows the center of mass of all the wheat production for each year. Each lat/lon pair is weighted based on the yield. From this graph, the center of mass does not appear to move North much, but surprisingly it does move East.

```{r vis_3, cache=TRUE, warning=FALSE}
  four_states = states %>% 
    filter(STUSPS == "IA"|
           STUSPS == "MO"|
           STUSPS == "KS" |
           STUSPS == "NE"
           ) %>% 
    st_crop(xmin = -96, xmax = -94,
    ymin = 39.2, ymax = 40.3)
  data %>% 
    filter(!is.na(yield), year != 1981) %>% 
    group_by(year) %>% 
    summarize(lat = weighted.mean(lat, w = yield, na.rm = TRUE), 
              lon = weighted.mean(lon, w = yield, na.rm = TRUE)) %>% 
    st_as_sf(coords=c("lon", "lat")) %>% 
    st_set_crs(value = 4326) %>%
    ggplot() + 
    geom_sf(data = four_states, fill = "white") + 
    geom_sf_text(aes(color = year, label = year), check_overlap = TRUE) +
    scale_color_gradient(low = "black", high = "orange") +
    xlab("") +
    ylab("") +
    labs( 
        title = "Center of Wheat Yield",
        subtitle = "Weighted average of centroid of wheat",
        color="Year")
```

## Visualization 4

To better capture the data from the previous visualization, I chose a different plot to represent the movement of the center of mass over time. There are two y-axis with latitude on the left and longitude on the right. 

It appears that the center of mass may be moving North after the year 2003, which is something I could dig into more. The Eastern trend is also shown, with less negative longitude values being more Easterly.

```{r viz_4, cache=FALSE, message=FALSE}

  data %>% 
    filter(!is.na(yield), year != 1981) %>% 
    group_by(year) %>% 
    summarize(lat = weighted.mean(lat, w = yield, na.rm = TRUE), 
              lon = weighted.mean(lon, w = yield, na.rm = TRUE)) %>% 
    pivot_longer(c("lat", "lon"), names_to = "type", values_to = "coord") %>%
    mutate(coord = ifelse(type == "lon", coord / -2.38, coord),
           type = ifelse(type == "lon", "Longitude", "Latitude")) %>% 
    ggplot(mapping = aes(x = year)) + 
    geom_point(mapping = aes(y = coord, color = type, shape = type), size=3) +
    geom_smooth(mapping = aes(y = coord, linetype = type, color = type)) +
    geom_line(mapping = aes(y = coord, linetype = type, color = type), alpha = 0.3) +
    scale_y_continuous(
      "Latitude", 
      sec.axis = sec_axis(~ . * -2.38, name = "Longitude")
    ) + 
    xlab("Year") + 
    labs( 
        title = "Shifting of Wheat",
        subtitle = "Weighted centroid of wheat") + 
    theme(legend.title=element_blank())
```

The next piece of data I would want to look at is if the lowest latitude and the highest latitude where wheat consistently grows has moved up or down, and compare that to the temperature graph that shows where it should grow. The center of mass may not be moving much because the bulk of the growth occurs in an environment that still meets the minimum or maximum temperature threshold to grow wheat. 

Technological advancements, rather than temperature changes, may allow for more production in the geographic areas which had sufficient temperature from 1981 to 2015. These advancements, or various other reasons, may account for center of gravity shifts more than temperature. This is why I want to look at some threshold value of production in the North and South rather than a weighted center next.


```{r, inclue=FALSE, echo=FALSE}
# Below is how to do this with raster data if the need arises

# array <- ncvar_get(data_2015, "var")
# fillvalue <- ncatt_get(data_2015, "var", "_FillValue")
# 
# array[array == fillvalue$value] <- NA
# 
# lat_2015 <- ncvar_get(data_2015, "lat")
# lon_2015 <- ncvar_get(data_2015, "lon")
# 
# r <- raster(t(array),
#   xmn = min(lon_2015), xmx = max(lon_2015),
#   ymn = min(lat_2015), ymx = max(lat_2015),
#   crs = CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs+ towgs84=0,0,0")
# )
# 
# r <- flip(r, direction = "y")
# plot(r)
# 
# data_1982 <- nc_open("gdhy_v1/wheat/yield_1982.nc4")
# 
# array <- ncvar_get(data_1982, "var")
# fillvalue <- ncatt_get(data_1982, "var", "_FillValue")
# 
# array[array == fillvalue$value] <- NA
# 
# lat_1982 <- ncvar_get(data_1982, "lat")
# lon_1982 <- ncvar_get(data_1982, "lon")
# 
# r <- raster(t(array),
#   xmn = min(lon_1982), xmx = max(lon_1982),
#   ymn = min(lat_1982), ymx = max(lat_1982),
#   crs = CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs+ towgs84=0,0,0")
# )
# 
# r <- flip(r, direction = "y")
# plot(r)

# The commented code below creates a wider table with one row per location and
# and a column for each year

# file_list <- list.files("data/gdhy_v1/wheat_spring/", pattern="nc4")
# 
# rm(wheat_data)
# 
# for(file_name in file_list) {
#   path <- paste("data/gdhy_v1/wheat/", file_name, sep="")
#   file <- tidync(path)
#   year <- parse_number(file_name)
#   
#   new_data <- file %>%
#     hyper_tibble() %>% 
#     rename("{year}":= var)
#   if (exists("wheat_data")){
#     wheat_data <- left_join(wheat_data, new_data)
#   } else {
#     wheat_data <- new_data
#   }
# }
```
